---
title: "data_analysis_one"
author: "Kade Stewart, Olivia Cowen, Kaitlyn Klein"
date: "2026-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this chunk we added and ran any necessary packages that we would need in order to begin our data analysis. These packages include the tidyverse package, the here package, and the car package. Tidyverse is the main package that we used for some of the major work with the data analysis. We used this to wrangle the data in a way that the different variables are able to be compared. The here package was used to simply show R where to find our data within our files. Lastly we have used the car package to further our data analysis and conduct various tests for normality within our data sets and and data analysis.

```{r}
library(tidyverse)
library(here)
library(car)
```

Once all of our packages were read in, we began the process of our data analysis by reading in the processed data from the data clean up. This was conducted through the "here" command. This command allowed R to more easily and accurately identify and load in our data.

```{r}
discharge_processed <- read_csv(here("data", "processed", "discharge_processed.csv"))
fish_processed <- read_csv(here("data", "processed", "fish_processed.csv"))
plant_processed <- read_csv(here("data", "processed", "plant_processed.csv"))
```

In this chunk we further cleaned our data to group the data points into various groups such as domain ID, the site ID, the year and month, as well as the fishes life stage within our fish data set. We also created new columns for monthly measurements. This further cleaning of our data allows for a smoother join of the three data sets for analysis.

```{r}
fish <- fish_processed %>% 
  group_by(domainID, siteID, year, month, fishLifeStage) %>% 
  summarise(fish_month_tally = n(),
            sum_fish_mass = sum(fishWeight, na.rm = TRUE))

plant <- plant_processed %>% 
  drop_na(aquaticPlantType) %>% 
  group_by(domainID, siteID, year, month) %>% 
  summarise(plant_month_tally = n())

discharge <- discharge_processed %>% 
  drop_na(dischargeContinuous) %>% 
  group_by(domainID, siteID, year, month) %>% 
  summarise(discharge_month_mean_L_sec = mean(dischargeContinuous),
            discharge_month_max = max(dischargeContinuous),
            discharge_month_min = min(dischargeContinuous))

```

The next important step for our data analysis was to be able to add the different data points from the different data sets together into one object so we could analyze the variables as functions of one another. In this section we also made the decision to abstain from using the plant data. This is due to the fact that the data set did not include as much information as orginally anticipated. However we have decided to keep it within our code in order to potentially revist at a later date. As for the Discharge and Fish data sets we joined them together by merging the columns domain ID, site ID, month, and year, as well as keeping the individual columns that were located in each data file.

```{r}
discharge_fish <- full_join(discharge, fish, by = c("domainID", "siteID", "month", "year"))

discharge_fish_plant <- full_join(discharge_fish, plant, by = c("domainID", "siteID", "month", "year"))
```

In this chunk we added in code in order to create a new object with new columns for the means of various metrics within our new data set. We also added in code to calculate the standard deviaton for the columns discharge_month_mean_L_sec and sum_fish_mass. These will both be important for the creation of error bars within the graphs created after the data analysis.

```{r}
disfish_site_means <- discharge_fish %>% 
  group_by(siteID) %>% 
  drop_na(discharge_month_mean_L_sec, sum_fish_mass) %>% 
  mutate(dis_mean = mean(discharge_month_mean_L_sec, na.rm = T),
         dis_sd = sd(discharge_month_mean_L_sec),
         fish_g_mean = mean(sum_fish_mass, na.rm = T),
         fish_g_sd = sd(sum_fish_mass))
```

One last piece of wrangling that was done to finalize our data for analysis was the removal of any NA values within our data set. After this, only numbers remain and should allow us to run the data analysis without running into issues due to the NA values.

```{r}
discharge_fish_na_rm <- discharge_fish %>% 
  drop_na(sum_fish_mass, discharge_month_mean_L_sec)
```

Create model #1 that compares average monthly fish mass to monthly discharge

```{r}
glm_fish_g_discharge <- glm(sum_fish_mass ~ discharge_month_mean_L_sec + siteID, data = discharge_fish)
summary(glm_fish_g_discharge)

#We need to ask how to do a Tukey on a glm because we cant see exactly what is significant...
```
Run diagnostic plots for model 1
```{r}
plot(glm_fish_g_discharge, which = 1:3)
shapiro.test(glm_fish_g_discharge$residuals)
```

Run model #2 that compares fish mass across sites

```{r}
anova_fish_g_site <- aov(sum_fish_mass ~ siteID, 
                  data = discharge_fish_na_rm)
summary(anova_fish_g_site)
TukeyHSD(anova_fish_g_site)
#No significance
```
Run diagnostic plots for model 2
```{r}
plot(anova_fish_g_site, which = 1:3)
leveneTest(anova_fish_g_site)
```

Run model #3 that compares monthly discharge across sites

```{r}
anova_discharge_site <- aov(discharge_month_mean_L_sec ~ siteID, 
                  data = discharge_fish_na_rm)
summary(anova_discharge_site)
TukeyHSD(anova_discharge_site)

#Some significance, only between MAYF and BIGC, TECR, and WALK
```
Run diagnostic plots for model 3
```{r}
plot(anova_discharge_site, which = 1:3)
leveneTest(anova_discharge_site)
```

For our statistical analysis we conducted both a linear regression test as well as two different ANOVA tests in order to gain an in depth analysis on how these variables were interacting with each other as well as how they differed among the different sites. 

For our first test we used the linear regression to test for the correlation of fish mass with montly discharge. For the linear regression we tested our residuals for normalcy with the shapiro test as well as tested our residuals vs fitted, Q-Q plot, and scale location. These graphs demonstated that our data was not normal. 

Next, we conducted two ANOVA test one testing monthly discharge across the various sites and the other to test for differences in fish mass across the various sites. We also conducted various test for these to test for normalcy such as the Levene's test, while also testing our assumptions using the residuals vs fitted, Q-Q, and scale-location plots. These graphs showed that again our data was not normal. 

However for the sake of gaining a general understanding of the data we proceeded with testing before greatly modifying our data. According to model one, only two sites showed a statistically significant correlation between fish mass and the monthly discharge site IDSYCA and site IDWALK with p-values of 0.007 and 0.006 respectively. 

For the first ANOVA test it can be seen that there is not statistically significant difference between fish mass at the various sites. 

However, in the second it could be seen that there was a statistically significant difference between MAYF-BIGC, TECR-MAYF, WALK-MAYF. The values portrayed by the tukey test show us that MAYF had the larger amount of monthly discharge among BIGC, TECR, and WALK. the p-values were as follow: 0.004, 0.004, 0.002 respectively.

Fish mass vs Discharge

```{r}
discharge_fish_na_rm %>% 
  ggplot(aes(x = discharge_month_mean_L_sec, y = sum_fish_mass, color = siteID)) + 
  geom_point() + 
  stat_smooth(method = "lm") +
  scale_y_continuous("Fish Mass (g)") +
  scale_x_continuous("Discharge (L/sec/month)")
```

The figure above shows the sum of all fish mass caught in a particular month at a particular NEON site (y-axis) compared to average monthly discharge at the same site and month (x-axis). Each color represents a different NEON site, and the lines that connect them are trendlines. There appears to be relatively positive correlation with WALK however the rest seem to appear relatively flat.

Mean fish mass across sites

```{r}
disfish_site_means %>% 
  ggplot(aes(x = siteID, y = fish_g_mean, colour = siteID)) +
  geom_bar(stat = "identity", 
           position = position_dodge2(preserve = "single")) +
  geom_errorbar(aes(ymin = fish_g_mean - fish_g_sd,
                    ymax = fish_g_mean + fish_g_sd),
                    position=position_dodge(.9),
                    width=.3, color = "black") +
  xlab("NEON Site") +
  ylab("Average Fish Mass per Month (g)") +
  theme_bw() 
```

The figure above shows the average fish mass per month at an individual site. The x-axis shows the NEON sites that are being examined, while the y-axis is the average fish mass in grams. Errors bars are calculated using the mean plus and minus standard deviation. As all error bars overlap can be seen that there is not a statistically significant difference between them. However, it would appear that SYCA and WALK have a lower general fish mass.

Mean discharge across sites

```{r}
disfish_site_means %>% 
  ggplot(aes(x = siteID, y = dis_mean, colour = siteID)) +
  geom_bar(stat = "identity", 
           position = position_dodge2(preserve = "single")) +
  geom_errorbar(aes(ymin = dis_mean - dis_sd,
                    ymax = dis_mean + dis_sd),
                    position=position_dodge(.9),
                    width=.3, color = "black") +
  xlab("NEON Site") +
  ylab("Average Discharge per Month (L/sec)") +
  theme_bw() 
```

The figure above shows the average discharge per month at an individual site. The x-axis shows the NEON sites that are being examined, while the y-axis is the average stream discharge in L/sec per month. Errors bars are calculated using the mean plus and minus standard deviation. Can be seen the TECR, WALK, and and BIGC have significantly lower average discharge than MAYF. Although not statistically significant, it would appear that discharge varies among the different sites.
