---
title: "NEON_data_cleanup"
output: html_document
date: "2026-02-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

 Load in Tidyverse and Here packages
```{r}
library(tidyverse)
library(here)
```

## Discharge data 

Read in the raw data for discharge and check structures of the variables. 
```{r}
discharge_raw <- read_csv(here("data", "raw", "csd_15_min.csv"))

head(discharge_raw)
str(discharge_raw)
#view(discharge_raw)

```

Begin wrangling our data by selecting different variables that could be important for future analysis across our different data tables. We also made sure to try and select variables that would be easy to refine and also variables that may be in common with the variables in different data sets. 

Add year and month for discharge from startDateTime; retain continuous stage and discharge data.
```{r}

discharge <- discharge_raw %>% 
  mutate(year = year(startDateTime),
         month = month(startDateTime)) %>% 
  select(domainID, siteID, year, month, stageContinuous, dischargeContinuous)

```

## Fish data 
Here we begun processing the fish data. First we needed to add in the raw data.
```{r}

fish_raw <- read_csv(here("data", "raw", "fsh_perFish.csv"))

head(fish_raw)
str(fish_raw)
#view(fish_raw)

```

Add year and month columns based on passStartTime.
Retain only adult lifestage; all samples retained are from electrofisher
Retain observations that have total length or weight

```{r}

fish <- fish_raw %>% 
  mutate(year = year(passStartTime),
         month = month(passStartTime)) %>% 
  filter(fishLifeStage == "adult") %>% 
  select(domainID, siteID, year, month,
         taxonID, taxonRank, scientificName,
         fishTotalLength, fishWeight) %>% 
  filter(!is.na(fishTotalLength) | !is.na(fishWeight))

```


here we read in the plant_raw data, and narrowed downt he plant variables we will be using to plot.
```{r}
plant_raw <- read_csv(here("data", "raw", "apc_perTaxon.csv"))

head(plant_raw)
str(plant_raw)
#view(plant_raw)

```
```{r}
plant <- plant_raw %>% 
  mutate(year = year(collectDate),
         month = month(collectDate)) %>% 
  select(domainID,
         siteID,
         namedLocation,
         collectDate,
         sampleCondition,
         aquaticPlantType,
         growthForm,
         taxonID,
         scientificName,
         morphospeciesID,
         phylum,
         division,
         class,
         order,
         family,
         genus,
         section,
         specificEpithet,
         infraspecificEpithet,
         variety,
         form,
         taxonRank,
         identifiedDate,
         dataQF)
  
```


Write a csv files for each processed data file: 
```{r}

discharge %>% 
  write_csv(here("data", "processed", "discharge_processed.csv"))

plant %>% 
  write_csv(here("data", "processed", "plant_processed.csv"))

fish %>% 
  write_csv(here("data", "processed", "fish_processed.csv"))

```


## Preliminary Plot
```{r}

ggplot(data = discharge_raw,
       aes(x = startDateTime, 
           y = dischargeContinuous,
           col = siteID)) +
  geom_point(cex = 1) +
  stat_smooth()

```
