---
title: "NEON_data_grab"
authors: Kade Stewart, 
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this chunk, we loaded the scripts 'tidyverse', 'here', and 'neonUtilities'. Tidyverse will be used to clean our data products, and the here package will be used to direct our markdown files to the proper working directory. NeonUtilities will be used to download and acess the NEON data that we will be using as the basis of our project. 
```{r library}
library(tidyverse)
library(here)
library(neonUtilities)
```

Hey this is a test on 2/17/2026!

We then pulled data from the NEON data product entitled 'Continuous Discharge' at all 7 sites we have selected to use in our analysis. With this data product, we plan to compare discharge between the west and east coast of the United States to see if there is a correlation between flow and fish abundance at each site.   
```{r}
#Pull data
discharge <- "DP4.00130.001"

#Select sites and dates *DATES SHOULD BE REEVALUATED
my_sites <- c("BIGC", "TECR", "REDB", "SYCA", "LECO", "WALK", "MAYF")
start_date <- "2018-11"
end_date <- "2022-12"

#Pull
list2env((loadByProduct(dpID = discharge, 
                        site = my_sites, 
                        include.provisional = FALSE,
                        startdate = start_date,
                        enddate = end_date)),
         .GlobalEnv)
```
Here, we wrote in our the variables that we will be using  
```{r}

#Save raw data DONE
variables_00130 %>% 
  write_csv(here("data", "raw", "variables_00130.csv"))

csd_15_min %>% 
  write_csv(here("data", "raw", "csd_15_min.csv"))

```
We pulled the NEON data product entitled "Fish electrofishing, gill netting, and fyke netting counts" at all 7 sites in order to get our 
```{r}
#Pull data from	"Fish electrofishing, gill netting, and fyke netting counts"
fish <- "DP1.20107.001"

#Select sites and dates 
my_sites <- c("BIGC", "TECR", "REDB", "SYCA", "LECO", "WALK", "MAYF")
start_date <- "2012-01"
end_date <- "2022-12"

#Pull
list2env((loadByProduct(dpID = fish, 
                        site = my_sites, 
                        include.provisional = FALSE,
                        startdate = start_date,
                        enddate = end_date)),
         .GlobalEnv)
```
Saving raw data variables 
```{r}
#Save raw data DONE
variables_20107 %>% 
  write_csv(here("data", "raw", "variables_20107.csv"))

fsh_perFish %>% 
  write_csv(here("data", "raw", "fsh_perFish.csv"))

```
Pulling NEON Data 'Aquatic plants' from from BIGC, TECR.
```{r}
#Pull data from "Aquatic plants"
plants <- "DP1.20072.001"

#Select sites and dates
my_sites <- c("BIGC", "TECR", "REDB", "SYCA", "LECO", "WALK", "MAYF")
start_date <- "2012-01"
end_date <- "2022-12"

list2env((loadByProduct(dpID = plants, 
                        site = my_sites, 
                        include.provisional = FALSE,
                        startdate = start_date,
                        enddate = end_date)),
         .GlobalEnv)   
```
Save raw variables 
```{r}
#Save raw data DONE
variables_20072 %>% 
  write_csv(here("data", "raw", "variables_20072.csv"))

apc_perTaxon %>% 
  write_csv(here("data", "raw", "apc_perTaxon.csv"))

```


```{r}
#Save raw data
variables_20086 %>% 
  write_csv(here("data", "raw", "variables_20086.csv"))

csd_15_min %>% 
  write_csv(here("data", "raw", "csd_15_min.csv"))
```