---
title: "NEON_data_grab"
authors: Kade Stewart, Olivia Cowen, Kaitlyn Klein 
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this chunk, we loaded the scripts 'tidyverse', 'here', and 'neonUtilities'. Tidyverse will be used to clean our data products, and the here package will be used to direct our markdown files to the proper working directory. NeonUtilities will be used to download and access the NEON data that we will be using as the basis of our project. 

```{r library}
library(tidyverse)
library(here)
library(neonUtilities)
```

# Set NEON data products, sites, dates
  
```{r}

discharge <- "DP4.00130.001"
fish <- "DP1.20107.001"
water_temp <- "DP1.20288.001"

start_date <- "2018-11"
end_date <- "2024-12"

# Set sites and dates 
my_sites <- c("BIGC", "TECR", "REDB", "SYCA", "LECO", "WALK", "MAYF")

```

## Download raw discharge data

We pulled data from the NEON data product entitled 'Continuous Discharge' at all 7 sites from 2018-2022. With this data product, we plan to compare discharge between the west and east coast of the United States to see if there is a correlation between flow and fish biomass at each site. 
```{r}

list2env((loadByProduct(dpID = discharge, 
                        site = my_sites, 
                        include.provisional = T)),
         .GlobalEnv)

```

Here, we wrote in the variables needed to begin data wrangling and analysis for discharge. The first csv is the metadata table which contains useful descriptions of the variables NEON provides in the data product. The second csv is the file containing our continuous discharge data.  
```{r}

# Save variables metadata
variables_00130 %>% 
  write_csv(here("data", "raw", "variables_00130.csv"))

# Save 15-min continuous discharge data
csd_15_min %>% 
  write_csv(here("data", "raw", "csd_15_min.csv"))

```

## Download raw fish data
We pulled the NEON data product "Fish electrofishing, gill netting, and fyke netting counts" for the same dates and sites. 

This data will allow us to find fish biomass and life stage data that we can use in our analysis.  

```{r}

list2env((loadByProduct(dpID = fish, 
                        site = my_sites, 
                        include.provisional = FALSE)),
         .GlobalEnv)
```

We saved the raw data from "Fish electrofishing, gill netting, and fyke netting counts" here. The first csv is the metadata table which contains useful descriptions of the variables NEON provides in the data product. The second csv file contains the fish taxon id and weight data which are potentially usefull variables for our data analysis. 
```{r}

# Save variables metadata
variables_20107 %>% 
  write_csv(here("data", "raw", "variables_20107.csv"))

# Save raw fish data
fsh_perFish %>% 
  write_csv(here("data", "raw", "fsh_perFish.csv"))

```

## Water temperature 
Not currently run/used
```{r}

list2env((loadByProduct(dpID = water_temp, 
                        site = my_sites, 
                        include.provisional = FALSE,
                        startdate = start_date,
                        enddate = end_date)),
         .GlobalEnv)   
```
