---
title: "NEON_data_cleanup"
Authors: Dr. F
output: html_document
---

## Import raw fish and discharge data

```{r, message = FALSE}
# Load packages
#install.packages("pacman")
pacman::p_load(tidyverse, here, car, rstatix, multcompView)

# Import raw data files 
discharge_raw <- read_csv(here("data", "raw", "csd_15_min.csv"))

fish_raw <- read_csv(here("data", "raw", "fsh_perFish.csv"))

```

## Discharge data wrangling

Goal: Summarize 15-minute discharge data into monthly summary values per site

```{r, message = FALSE}

discharge <- discharge_raw %>% 
  mutate(year = year(startDateTime),
         month = month(startDateTime)) %>% 
  drop_na(dischargeContinuous) %>% 
  group_by(domainID, siteID, year, month) %>% 
  summarise(discharge_month_mean_L_sec = mean(dischargeContinuous),
            discharge_month_max = max(dischargeContinuous),
            discharge_month_min = min(dischargeContinuous))

```

## Fish data wrangling

Goals:

-   Keep fish that are ID'd to species/subspecies level.

-   Keep life stages that are big enough to measure (juvenile, gravid female, adult).

-   Exclude individuals with missing body mass.

```{r}

fish <- fish_raw %>% 
  mutate(year = year(passStartTime),
         month = month(passStartTime)) %>%  
  select(domainID, siteID, year, month,
         taxonID, taxonRank, fishLifeStage, scientificName,
         fishWeight) %>% 
  filter(!is.na(fishWeight),
         taxonRank %in% c("species", "subspecies"),
         fishLifeStage %in% c("adult", "juvenile", "gravid female"))

```

Calculate monthly site-level metrics for fish:

-   Abundance (total individuals)

-   Species richness (number of unique species)

-   Mean biomass (g)

-   Total biomass (sum; g)

```{r, message = FALSE}

fish_monthly <- fish %>% 
  group_by(domainID, siteID, year, month) %>% 
  summarize(abundance = n(),
            richness  = n_distinct(taxonID), 
            mean_biomass_g = mean(fishWeight, na.rm = T),
            total_biomass_g = sum(fishWeight, na.rm = T))

head(fish_monthly)

```

## Join monthly discharge and fish
Use inner_join to only retain complete cases (discharge + fish data)

```{r, message = FALSE}

monthly_data <- inner_join(discharge, fish_monthly)

head(monthly_data)

```

## Compare fish and discharge metrics among sites

### Monthly fish

```{r}

fish_mean_biomass_aov <- aov(mean_biomass_g ~ siteID, data = fish_monthly)

par(mfrow = c(1, 3))

plot(fish_mean_biomass_aov, which = c(1:3))

shapiro.test(fish_mean_biomass_aov$residuals)

leveneTest(mean_biomass_g ~ siteID, data = fish_monthly)

```
Mean biomass data do not meet assumption for normality or homogeneity of variance

```{r}

# Log transform data to improve normality
fish_monthly <- fish_monthly %>% 
  mutate(log_mean_biomass = log(mean_biomass_g))

# Welch ANOVA (for non-homogeneous data)
welch <- oneway.test(log_mean_biomass ~ siteID, data = fish_monthly, 
            var.equal = FALSE)

print(welch)

# Post-hoc test
biomass_post_hoc <- fish_monthly %>%
  ungroup() %>% 
  games_howell_test(log_mean_biomass ~ siteID)

```


```{r}

# Asssign letter groups based on sites that are sig. diff.
pvals <- biomass_post_hoc %>%
  mutate(comparison = paste(group1, group2, sep = "-")) %>%
  select(comparison, p.adj) %>%
  tibble::deframe()

letters <- multcompView::multcompLetters(pvals)$Letters

letters_df <- tibble(siteID = names(letters),
                     letters = letters) %>%
  arrange(siteID)

```

Plot of these data
```{r}

# Force siteID to list alphabetically
fish_monthly_plot <- fish_monthly %>% 
  mutate(siteID = factor(siteID, levels = c("BIGC", "LECO", "MAYF",
                                            "REDB", "SYCA", "TECR", "WALK")))

# Plot! 
ggplot(fish_monthly_plot, 
       aes(x = siteID, 
           y = mean_biomass_g,
           col = siteID)) +
  geom_jitter(width = 0.1, height = 0,
              size = 3, alpha = 0.5) +
  geom_boxplot(fill = NA, col = "black") +
  geom_text(data = label_pos,
            aes(x = siteID, y = exp(y) + 1, label = letters),
            vjust = 0, color = "black") +
  theme_minimal()
```

## Assess fish ~ discharge relationships

### Fish richness = count data, need to use poisson regression

```{r}
mod <- glm(richness ~ discharge_month_mean_L_sec, 
           data = monthly_data, 
           family = poisson)

# Check for overdisperson
mod$deviance / mod$df.residual

# Overdispersed! Pivot to negative binomial model
mod_nb <- glm.nb(richness ~ discharge_month_mean_L_sec, 
           data = monthly_data)

summary(mod_nb)

```

```{r}

ggplot(data = monthly_data,
       aes(x = discharge_month_mean_L_sec,
           y = richness,
           col = siteID)) +
  geom_point(size = 3, alpha = 0.5) +
  geom_line(aes(y = fitted(mod_nb), group = 1), 
            color = "black") +
  theme_minimal()

```

